---
title: "Bayesian Estimation of Monotone Single Index Models"
author: "Snigdha Das"
output: 
  github_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The package `monotoneSIM` performs Bayesian estimation using B-Spline basis approximation of a Single Index Model where the unknown link function is assumed to be monotonically increasing. 
Consider a Single index models of the form: 
$$y = g(x^{\top} \beta) + \epsilon$$

where $y$: response, $x$: p-dimensional predictors, $\beta$: p-dimensional coefficient vector,  $\epsilon$: independent N($0,\ \sigma_\epsilon^2$) random errors, $g$: unknown monotone link function. 
To construct a prior on $g$, we consider the basis expansion: $g(x) = \Sigma^{L}_{l=0} \, \xi_l \, \psi_l(x)$, where $\psi_l(x) = \int^{x}_{-1} h_l(t)\,\text{d}t$ , $h_l$ is a B-Spline basis function of order 2. Then, $g$ is monotonically increasing if and only if $\xi_l \geq 0 \ \forall \ l$, thereby enforcing the monotonicity constraint in an equivalent way. 

Note: Here $g(x) = 0$, when $x \le -1$ and $g(x) = g(1)$, when $x \ge 1$. The matrix of predictors is scaled so that every row has euclidean norm $\le 1$ and euclidean norm of $\beta$ is taken to be 1 so that $|x^\top \beta| \le 1$.

## Functionality

The package provides two functions.

* The main function `monotoneSIM` performs a Markov Chain Monte Carlo algorithm to generate samples from the conditional posterior distribution of unknown parameters in a monotone Single Index Model - the unknown parameters being B-Spline basis coefficients that approximate the unknown link function, the single index parameter and the error variance of the model.

* The other function `monotoneFIT` calculates fitted values of the response using posterior mean of the single index parameter after generating an MCMC sample using the function `monotoneSIM`. It also returns the estimated link function of the model based on basis coefficients for a grid of supplied values.

## Installation

To install this package from Github, run the following in your R console:
```{r echo = TRUE, eval = FALSE}
devtools::install_github("das-snigdha/monotoneSIM")
```

## An Example Demonstrating Usage 

This package shall be implemented on the following simulated data.

We take two continuous variables and one dichotomous attribute as predictors.
```{r echo = TRUE, eval = TRUE}
n = 100; p = 3;
X = matrix(rnorm(n*(p-1)), nrow = n, ncol = (p-1))
X = cbind(X, rbinom(n, 1, 0.5))

```

Let the true monotone link function be : $$ g(x) = 5 \bigg[ F((x+1)/2) - F(0) \bigg] ,-1 \le x \le 1.$$

```{r echo = TRUE, eval = FALSE}
# True monotone link function
true.g = function(x){
  y = (x+1)/2
  5*(pnorm(y, mean=0.5, sd=0.1) - pnorm(0, mean = 0.5, sd = 0.1))
}
plot(true.g, xlim = c(-1,1))

```

Clearly g(.) is monotone increasing and $g(-1) = 0$.

The following tasks shall be performed during the remaining of the semester:

* Perform Compatibility Checks on user supplied input.
* Select equispaced `knots`, $u_0, u_1, \dots, u_L$ between -1 and 1, if not supplied.
* Select grid of x values, `grid.x` between -1 and 1 of length `size.grid.x`, if not supplied
* Perform scaling of covariate matrix `X` and supplied `beta.init`.
* Perform back-scaling on obtained `beta`, to get coefficients corresponding to unscaled X.
* Create a vignette to demonstrate the usage, if time permits.


